---
title: 'StatBI/E Exercises: Week 6'
output:
  html_notebook:
    code_folding: none
    toc: yes
    toc_float:
      collapsed: no
    df_print: paged
    theme: sandstone
    highlight: kate
  html_document:
    toc: yes
    df_print: paged
  pdf_document:
    toc: yes
---
<style>
.exercise{background-color:#c9e2a5;padding:1ex 1ex .1ex 1ex;}
.answer{background-color:#f3f3a5;padding:1ex 1ex .1ex 1ex;}
</style>
::: exercise :::
[This is how exercises are highlighted]
:::
::: answer :::

:::
</small>

# Exercise 1: Beer foam

In this problem we analyze the `beerfoam` data in the `week6.RData` file, 
which you can load via
``` {r}
load("week6.RData")
# the variable beerfoam then holds the data
head(beerfoam)
```
The data set contains 13 observations of measurements of wet foam
height and beer height at various time
points for Shiner Bock at 20C.
There are three variables:

* t: the time from pour (in seconds)
* foam: the wet foam height (cm)
* beer: the beer height (cm)

We want to estimate how the height of the foam decreases over time.



## Exercise 1.1

::: exercise :::
Fit the simple linear regression
$$\mathbb{E}(\text{foam}|t) = β_0 + β_1 t$$
Plot the observations $(t, \text{foam})$ and the linear regression obtained.
Do you think the model is appropriate?
(Hint: plot t versus the residuals)
:::

::: answer :::

:::



## Exercise 1.2

::: exercise :::
Fit a quadratic regression
$$\mathbb{E}(\text{foam}|t) = β_0 + β_1t + β_2 t^2$$
Plot the observed points and the fitted quadratic regression, what you can
observe?
Does the model appear to fit the data better than the simple linear regression above?

Perform the F-test and the model selection using AIC and BIC between the simple linear
regression (Ex. 2.1) and the quadratic regression (Ex. 2.2).
What does the quadratic model predict as $t$ increases (e.g. $t = 600$
seconds)?
Does the prediction of the model make sense?
Explain and comment
(remember the physical experiment we are analyzing).
:::

::: answer :::

:::



## Exercise 1.3

::: exercise :::
Fit the simple linear regression for $\log(\text{foam})$, this can
be done since foam is always a positive variable:
$$
\mathbb{E}(\log(\text{foam})|t) = β_0 + β_1t$$
(this is a simple linear regression over a transformed variable, it is not a
generalized linear regression).

Plot the regression function on top of the observed points and perform model
selection using AIC and BIC for all three models obtained in Ex 1.1–1.3 models (SL, Q and E). 
:::

::: answer :::

:::



## Exercise 1.4

::: exercise :::
We wish to investigate the relationship between foam and
beer; that is, the relationship between the height of the foam and the height of
the beer.
Try to fit at least $3$ different types of linear regression (and generalized linear
regression models) and comment on the results, you can use AIC and BIC to
preform model selection. 
(Hint: plot the observations first and try to guess possible models)

:::

::: answer :::

:::



# Exercise 2: Tooth Growth

In this problem we analyze the `ToothGrowth` data in the `week6.RData` file, 
which you can load via
``` {r}
load("week6.RData")
# the variable ToothGrowth then holds the data
head(ToothGrowth)
```

The data is used to examine the effect of
vitamin C on tooth growth in Guinea pigs.
The length of odontoblasts (cells responsible for
tooth growth) were recorded in 60 Guinea pigs. Each animal received one of
three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery
methods: orange juice (OJ) and ascorbic acid (VC). The data frame contains
three columns:

* len: tooth length
* supp: supplement type (VC or OJ)
* dose: dose level



## Exercise 2.1

::: exercise :::
Compute the mean tooth length for all six combinations of
supplement types and levels. Also provide the standard error of the mean for
each situation.
:::

::: answer :::

:::



## Exercise 2.2

::: exercise :::
We will investigate whether different dose levels have the same
effect. Perform 0.05-level two sample t-tests with (unknown) unequal variances to check
whether to reject the following null hypotheses, and explain the result for each
hypothesis:

* With the OJ method, the dose levels 0.5 and 1.0 mg/day come with the same
resulting tooth length.
* With the OJ method, the dose levels 1.0 and 2.0 mg/day come with the same
resulting tooth length.
* With the VC method, the dose levels 0.5 and 1.0 mg/day come with the same
resulting tooth length.
* With the VC method, the dose levels 1.0 and 2.0 mg/day come with the same
resulting tooth length.
:::

::: answer :::

:::


## Exercise 2.3

::: exercise :::

We are interested in whether OJ is more effective than VC.
Perform 0.05-level two sample t-tests with (unknown) unequal variances to check whether
to reject the following null hypotheses:

* With 0.5 mg/day dose level, OJ is less effective than or as effective as VC
in tooth growth.
* With 1.0 mg/day dose level, OJ is less effective than or as effective as VC
in tooth growth.
* With 2.0 mg/day dose level, OJ is less effective than or as effective as VC
in tooth growth.

Under which dose level(s) can we say OJ is more effective than VC?

:::

::: answer :::

:::



# Optional Bonus Exercise 3: A case study of neuronal data (continuation of Week 5, Exercise 1)

In this problem we analyze the `isi` data in the `week6.RData` file, 
which you can load via
``` {r}
load("week6.RData")
# the variable isi then holds the data
head(isi)
```

As part of previous exercises we have described and analysed the data.
Here, we perform Bayesian inference on the rate parameter of the exponential distribution.

For this, we assume the `isi` observations are
i.i.d. observations from an exponential distribution
with unkown rate parameter $\lambda > 0$.

As prior distribution we assume a Gamma distribution over $\lambda$, that is,
$\lambda \sim \operatorname{Gamma}(1, 1)$
where $\operatorname{Gamma}(\alpha,\beta)$ is
the Gamma distribution with shape parameter $\alpha$
and rate parameter $\beta$.



## Exercise 3.1

::: exercise :::
Implement the unnormalized log-posterior density and
obtain the MAP estimator for $\lambda$
using numerical optimization and the function `optimize`.
:::

::: answer :::

:::



## Exercise 3.2

::: exercise :::
The Gamma distribution is a conjugate prior of the exponential model and
thus closed form expression for the posterior is known, in particular if
$\lambda\sim \operatorname{Gamma}(\alpha, \beta)$
and
$X_i \sim \operatorname{Exponential}(\lambda)$
then the posterior distribution of $\lambda$ is
once again a Gamma distribution with parameters

$$\text{shape} = \alpha + n$$
and
$$\text{rate} = \beta + \sum_{i=1}^n X_i$$

Implement the closed form expression of the
posterior mean estimation for $\lambda$
and compare it to the MAP estimation from above.

Hint:
If $X \sim \operatorname{Gamma}(\alpha, \gamma)$
then
$\mathbb{E}(X) = \frac{\alpha}{\beta}$
and
$\mathbb{V}(X) = \frac{\alpha}{\beta^2}$.
It may also be helpful
to plot the density of the Gamma distribution for the different choices of the
parameters.
:::


::: answer :::

:::


\