---
title: "expectation and variance"
author: ""
date: ''
output:
  html_notebook:
    code_folding: none
    toc: yes
    toc_float:
      collapsed: no
    df_print: paged
    theme: sandstone
    highlight: kate
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---
<style>
.exercise{background-color:#c9e2a5;padding:1ex 1ex .1ex 1ex;}
.answer{background-color:#f3f3a5;padding:1ex 1ex .1ex 1ex;}
</style>
::: exercise :::
[This is how exercises are highlighted]
:::
::: answer  :::
[This is how your answers are highlighted]
:::




## Numerical integration 

To perform numerical integration of a one-dimensional function,
we can use the  `integrate` function.

For example, we can compute $\int_{2}^5 x^2 + \cos(x) dx$ as follows
``` {r}
f <- function(x) {
   return(x ^ 2 + cos(x))
}
integrate(f, lower = 2, upper = 5)
```

The function `integrate` accepts the following parameters:

- `f` the function to integrate, must accept a numerical vector and return a
   numerical vector of the same length;
- `lower` and `upper` are the lower and upper limits of integration, respectively,
they can be `-Inf` or `Inf`
- there are further optional parameters, cf. `?integrate`

It is a requirement that the function `f` 
accepts a numerical vector and returns a numerical vector of the same length.
For example, the following fails
``` {r error=T}
f <- function() {
   return(1)
}
integrate(f, lower  = 0, upper = 1)
```

::: exercise ::: 
@. Describe the problem in above code chunk.
:::


::: answer :::

:::


Also, the following fails
because the function `f` does not return a vector
``` {r error=T}
f <- function(x) {
   return(1)
}
integrate(f, lower  = 0, upper = 1)
```
To make the example work, we can change the example as follows
``` {r}
f <- function(x) {
   return(rep(1, length(x)))
}
integrate(f, lower  = 0, upper = 1)
```

Instead, we could have used the function `Vectorize`
to _vectorize_ the original function definition;
for example
``` {r}
g <- function(x) {
   return(1)
}

g_vect <- Vectorize(g)

integrate(g_vect, lower = 0, upper = 1)
```

::: exercise :::
@. Compute $\int_{-\pi}^{\pi} cos(x)sin(x)\,dx$ 
@. Compute $\int_{0}^{10} \min\{x, 4\}\, dx$
@. Numerically verify that the functions `dnorm`, `dgamma`, and `dexp`
   are probability density functions, that is,
   that they integrate to 1 and are non-negative.
   Do this for different settings of the parameters.
@. Define a new probability density function `dsin`
   of a random variable $X$ such that
   $f_X(x) \propto \sin(x)$ for $x \in (0,\pi)$ and $0$ otherwise.
   Visually inspect, that the function is non-negative
   and compute the normalization constant using `integrate`.
:::

::: answer :::

:::

## Expectations of random variables

We can use numerical integration to compute the expected value of 
random variables and of functions of random variables
(using the _rule of the lazy statistician_).

For example, we can obtain the expected value of  $X^2 + \sin(X)$
where $X \sim N(3,.5^2)$
as follows
``` {r}
m = 3
s = 0.5
xf <- function(x) {
   return((x ^ 2 + sin(x)) * dnorm(x, mean = m, sd = s))
}
integrate(xf, lower = -Inf, upper = Inf)
```

## Variance

We can obtain the variance of a random variable using the
formula
$$ \mathbb{V}(X) = \mathbb{E}(X^2) - \mathbb{E}(X)^2$$

We can, for example, obtain the variance of a Gaussian random variable 
$X \sim N(3, 4^2)$ as
``` {r}
f <- function(x) {
   return(dnorm(x, mean = 3, sd = 4))
}
xf <- function(x) {
   return(x * f(x))
}
xxf <- function(x) {
   return(x ^ 2 * f(x))
}
var_x <- integrate(xxf, lower = -Inf, upper = Inf)$value -
   integrate(xf, lower = -Inf, upper = Inf)$value ^ 2
var_x 
```

::: exercise ::: 
@. Write a generic function that takes as input the density function and returns 
the variance. 
@. Obtain the variance of $Y= \sin(X)$ where $X \sim Beta(\mathrm{shape1} = 5,\ \mathrm{shape2} = 4)$.
:::

::: answer :::

:::

## Empirical mean 

To compute the empirical mean $\bar{x} = \frac{\sum_{i=1}^n x_i}{n}$ we use
the `mean` function. It computes the arithmetic mean of a vector.
``` {r}
x_obs <- runif(100, min = 0, max = 10)
mean(x_obs)
```

We can also exclude `NA` values with the `na.rm = TRUE` argument,
``` {r}
obs <- c(13, 0.5, NA, 34, 9, NA)
mean(obs, na.rm = TRUE)
```

::: exercise :::
@. Visually compare the empirical mean and the true mean of a Gaussian random
variable for different values of the variance parameter and for an increasing 
number of observations (e.g. $n = 10,100,1000, 5000, 10000$).
@. Sample $n=100$ i.i.d. observations from an exponential distribution with 
 $\lambda = 1$ (rate) and compute the empirical mean. Repeat the experiment 100
 times and record the 100 computed empirical means. 
:::

::: answer :::

:::

## Empirical variance and standard deviation

The empirical variance $s^2 = \sum_{i=1}^n (\bar{x} - x_i)^2 / (n-1)$ can be 
obtained using the `var` function that takes as input a vector of 
observations.
``` {r}
sample <- rnorm(100)
var(sample)
```

As for the `mean` function, we can exclude `NA` values,
``` {r}
var(c(3, 2, 12, 4, NA, 54), na.rm = TRUE)
```

::: exercise :::
@. Sample $100$ observations from a normal random variable $X$ and compute its 
empirical variance, then compute the transformed data under the 
scale-location transformation $Y = 2X + 1$ and obtain the empirical 
variance of $Y$. 
@. Normalize the observations obtained in the previous exercise.
Verify that the empirical mean of the normalized data is $0$ and
the empirical variance of the normalized data is $1$.
@. Create a `data.frame` of $100$ observations of $100$ Gaussian random variables 
   $X_i$  
   with means $\mu = 1$ and unit variance. 
   For every variable $X_i$ obtain the empirical mean $\bar{X}_i$ 
   and the empirical variance.
   Then, for every empirical mean $\bar{X}_i$ 
   calculate the standard error of the mean.
   Compare the standard errors of the mean with 
   the empirical standard deviation of the $100$ 
   empirical means $\bar{X}_1, \bar{X}_2, \ldots, \bar{X}_{100}$. 
:::

::: answer :::

:::
